Toronto's homeless death toll over the years shows the dangers of addiction's epidemic

### 2024-01-25 outputs/paper/paper.qmd
adding files and finished citation

---
title: "How Toronto's homeless death toll over the years shows the dangers of addiction's epidemic?"
subtitle: "Data analysis of homeless deaths in Toronto from 2017 to 2023"
author: 
  - Shipeng Zhang
thanks: "Code and data are available at: LINK."
date: 2024-01-22
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
---

@@

You can and should cross-reference sections and sub-sections.

The remainder of this paper is structured as follows. @sec-data....

@@

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| warning: false
#| message: false
#| echo: false

file_path <- "../../inputs/data/raw_data.csv"
toronto_homeless_death <- read.csv(file_path)
toronto_homeless_death <- dplyr::select(toronto_homeless_death, -X_id)           
sample_data <- head(toronto_homeless_death, 6)
knitr::kable(sample_data,'markdown',col.names = c("Year of Death", "Cause of Death", "Age Group", "Gender", "Count"),align = 'c',

                      caption = "Sample Table for Causes of Death")
  # Print the table

@@

```{r}
#| label: fig-causes-of-death
#| fig-cap: Causes of death analysis
#| echo: false

ggplot(data = toronto_homeless_death,mapping =  aes(x = Cause_of_death, fill = Age_group)) +
  geom_bar(position = "stack") +
  scale_fill_manual(values = c("lightblue", "lightgreen", "lightpink", "lightyellow", "lightgray")) +
  labs(title = "Causes of Death Stacked by Age Group",
       x = "Cause of Death",
       y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but
don't worry about doing that until you have finished every other aspect
of the paper - Quarto will try to make it look nice and the defaults
usually work well once you have enough text.)

#\| label: fig-planes #\| fig-cap: Relationship between wing length and
width #\| echo: false #\| warning: false #\| message: false

analysis_data \<- read_csv(here::here("outputs/data/analysis_data.csv"))

analysis_data \|\> ggplot(aes(x = width, y = length)) + geom_point(alpha
= 0.8) + theme_minimal() + labs(x = "Wing width (mm)", y = "Wing length
(mm)")

```{r}

@@

Talk way more about it.

# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to
investigate... Background details and diagnostics are included in
[Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft.
Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both
measured in millimeters.

```{=tex}
\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}
```
We run the model in R [@citeR] using the `rstanarm` package of
@rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time
spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance
$\theta$.

Our results are summarized in @tbl-modelresults. #\| echo: false #\|
eval: true #\| warning: false #\| message: false

library(rstanarm)

first_model \<- readRDS(file =
here::here("outputs/models/first_model.rds"))

```{r}

#\| echo: false #\| eval: true #\| label: tbl-modelresults #\| tbl-cap:
"Explanatory models of flight time based on wing width and wing length"
#\| warning: false

modelsummary::modelsummary( list( "First model" = first_model ),
statistic = "mad", fmt = 2 )

```{r}

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The
discussion is a chance to show off what you know and what you learnt
from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\appendix

# Appendix {.unnumbered}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive
check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the
prior. This shows... #\| eval: true #\| echo: false #\| message: false
#\| warning: false #\| label: fig-ppcheckandposteriorvsprior #\|
layout-ncol: 2 #\| fig-cap: "Examining how the model fits, and is
affected by, the data" #\| fig-subcap: \["Posterior prediction check",
"Comparing the posterior with the prior"\]

pp_check(first_model) + theme_classic() + theme(legend.position =
"bottom")

posterior_vs_prior(first_model) + theme_minimal() +
scale_color_brewer(palette = "Set1") + theme(legend.position =
"bottom") + coord_flip()

```{r}

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests... #\|
echo: false #\| eval: true #\| message: false #\| warning: false #\|
label: fig-stanareyouokay #\| fig-cap: "Checking the convergence of the
MCMC algorithm" #\| fig-subcap: \["Trace plot", "Rhat plot"\] #\|
layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")

```{r}

# References

>>>>

---
title: "Toronto's homeless death toll over the years shows the dangers of addiction's epidemic"
subtitle: "Data analysis of homeless deaths in Toronto from 2017 to 2023"
author: 
  - Shipeng Zhang
thanks: "Code and data are available at: https://github.com/zero616/Homeless_Death_Analysis."
date: 2024-01-22
date-format: long
abstract: "Homelessness is a well-known problem in Toronto, and this article takes an in-depth look at the tragic reality of homelessness in Toronto. Focuses on the devastating impact of drug addiction by analyzing homeless death toll data from 2017 to 2023. Our top findings reveal a worrying rise in addiction-related homeless deaths over the years. By examining causes of death, death year and ratio of drug caused death, we provide a comprehensive overview revealing the dangers faced by this vulnerable group."
format: pdf
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
---

@@

Toronto is currently experiencing a growing drug toxicity crisis, with
challenges posed by the Covid-19 pandemic further exacerbating the
public health dilemma. This article explores the details behind the
crisis through a comprehensive analysis of homeless deaths from 2017 to
2023.

Drug addiction has a profound impact on the homeless community. Our
results identify a rapid expansion of drug toxicity deaths in Toronto.
This article takes an in-depth look at the proportion of deaths caused
by drug toxicity and reveals the impact of the COVID-19 pandemic on the
drug toxicity crisis.

Our article is divided into different sections: data, results,
discussion and conclusion. The Data section clarifies the nature of the
data-set obtained through the City of Toronto's OpenDataToronto (Toronto
Public Health 2023) and the tables that contributed to our analytic
work. In the results section, we show the clear growing trends that
emerged during the analysis. The Discussion section assesses these
trends and provides some insights into the underlying factors that led
to the crisis and its exacerbation by the pandemic. Finally, the
conclusion provides a concise summary based on the main findings,
outlining the seriousness of the crisis and the need for targeted
interventions.

The purpose of this article is to summarize the key challenges facing
homeless people in Toronto. Through detailed analysis of the data, we
deepen our understanding of this crisis and help inform data-based
policies and interventions.

@@

The data set used in this analysis was obtained from the City of
Toronto's OpenDataToronto Library (Gelfand 2022), adhering to ethical
guidelines and ensuring personal anonymity. It includes important
information such as year of death, cause of death, age group, gender and
number of people. For privacy and ethical reasons, the data does not
include names, detailed time of death, and place of death. The data
comes from data released by official coroner's agencies and shelters,
among others. To maintain transparency and reproducibility, all analyzes
were performed using the R programming language and visualized with
ggplot2 (Wickham 2016) and tables created with knit (Xie 2023).

@tbl-Death provides a summary of the data set outlining the key
variables under consideration: time to death, cause of death, sex and
age group. The time of death started in 2017 and ended in 2023, and the
age groups were divided into \<20, 20-39, 40-59, 60+.

```{r}
#| label: tbl-Death
#| tbl-cap: Toronto Homeless Death (2017-2023)
#| warning: false
#| message: false
#| echo: false
#read file
file_path <- "../../outputs/data/cleaned_data.csv"
toronto_homeless_death <- read.csv(file_path)
sample_data <- head(toronto_homeless_death, 6)
# Print the table
knitr::kable(sample_data,'markdown',col.names = c("Year of Death", "Cause of Death", "Age Group", "Gender", "Count"),align = 'c',

                      caption = "Sample Table of Toronto Homeless Death (2017-2023)")

@@

@tbl-Death-by-age provides the total number of deaths summarized by age
group and cause of death in the data set, outlining the key variables
available for analysis: cause of death, age group and number of deaths.

```{r}
#| label: tbl-Death-by-age
#| tbl-cap: Toronto Homeless Death(by age) (2017-2023)
#| warning: false
#| message: false
#| echo: false
# group by age
total_count_age_data <- toronto_homeless_death %>%
  group_by(Age_group, Cause_of_death) %>%
  summarise(total_count_age = sum(Count), .groups = 'drop')
sample_total_count_by_age_data <- head(total_count_age_data, 6)
knitr::kable(sample_total_count_by_age_data,'markdown',col.names = c("Age Group", "Cause of Death", "Total Count"),align = 'c',

                      caption = "Sample Table for total death number of causes of death by Age Group (2017-2023)")
```

@tbl-Death-by-year provides the total number of deaths in the data set
summarized by year of death and cause of death, outlining the key
variables in the analysis of addiction harm to homeless people: cause of
death, time of death, and number of deaths.

```{r}
#| label: tbl-Death-by-year
#| tbl-cap: Toronto Homeless Death(by year) (2017-2023)
#| warning: false
#| message: false
#| echo: false

#group by year
total_count_data <- toronto_homeless_death %>%
  group_by(Year.of.death, Cause_of_death) %>%
  summarise(total_count = sum(Count), .groups = 'drop')
sample_total_count_by_year_data <- head(total_count_data, 6)
knitr::kable(sample_total_count_by_year_data,'markdown',col.names = c("Year of Death", "Cause of Death", "Total Count"),align = 'c',

                      caption = "Sample Table for total death number of causes of death by Year (2017-2023)")
```

@fig-causes-of-death-by-age shows the distribution of causes of death
among homeless people, stratified by age group. This figure provides an
impact sketch of the leading causes of death among homeless people.

```{r}
#| label: fig-causes-of-death-by-age
#| fig-cap: Causes of death analysis by abe
#| warning: false
#| message: false
#| echo: false

#print figure
ggplot(data = total_count_age_data,mapping =  aes(x = Cause_of_death,y =total_count_age, fill = Age_group)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("lightblue", "lightgreen", "lightpink", "lightyellow", "lightgray")) +
  labs(title = "Causes of Death Stacked by Age Group (2017-2023)",
       x = "Cause of Death",
       y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

@fig-causes-of-death-by-year shows the number of homeless deaths
(stratified by cause of death) starting in 2017 and ending in 2023. This
graphic provides a visual representation of how the homeless drug
intoxicating crisis is developing. From Figure 2 we can see that
starting in 2020, the number of deaths began to increase rapidly and
peaked in 2021.

```{r}
#| label: fig-causes-of-death-by-year
#| fig-cap: Causes of death analysis by year
#| echo: false
#| warning: false
#| message: false
library(ggplot2)
library(dplyr)

# Create a stacked bar chart
ggplot(data = total_count_data, aes(x = Year.of.death, y = total_count, fill = Cause_of_death)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c( "#cc340c","#e8490f","#f18800","#e4ce00",
            "#9ec417","#13a983","lightgreen", "lightblue","#44c1f0","#3f60aa")) +
  labs(title = "Total Count of Causes of Death Stacked by Year (2017-2023)",
       x = "Year of Death",
       y = "Death Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0))
```

Even though we can clearly see the huge proportion of deaths caused by
drug toxicity and the worsening status quo from Figure 2, in order to
understand the worsening of the crisis more intuitively from the data, I
calculated and summarized the number of drug toxicity cause death as a
proportion of the total deaths The ratio of the number of people has
been compiled into the able and the figure, as well as @tbl-Death-ratio
and Figure 3.

```{r}
#| label: tbl-Death-ratio
#| tbl-cap:  ratio of Deaths Caused by Drug Toxicity (2017-2023)
#| warning: false
#| message: false
#| echo: false
# Calculate the ratio (percentage) of Drug Toxicity deaths for each year
ratio_data <- total_count_data %>%
  group_by(Year.of.death) %>%
  summarise(total_deaths = sum(total_count),
            drug_toxicity_deaths = sum(total_count[Cause_of_death == "Drug Toxicity"]),
            ratio = (round((drug_toxicity_deaths / total_deaths) * 100,digits=2)))  # Multiply by 100 to convert to percentage and round to .00
knitr::kable(ratio_data,'markdown',col.names = c("Year of Death", "Total Death", "Drug Toxicity Death", "Ratio"),align = 'c',

                      caption = "Sample Table for Ratio of Deaths Caused by Drug Toxicity Over Time (2017-2023)")
```

```{r}
#| label: fig-ratio-of-death
#| fig-cap: ratio of death Caused by Drug Toxicity Over Time
#| warning: false
#| message: false
#| echo: false
# Create a line chart for the ratio of Drug Toxicity deaths over time
ggplot(data = ratio_data, aes(x = Year.of.death, y = ratio)) +
  geom_line(color = "#f46f20",linewidth = 1.5) +
  geom_text(aes(label = sprintf("%.2f%%", ratio)), vjust = -0.5, hjust = 0.5, size = 3) +
  labs(title = "Ratio of Deaths Caused by Drug Toxicity Over Time (2017-2023)",
       x = "Year of Death",
       y = "Ratio (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0))
```

## Cause of death: {#sec-first-point}

Our analysis takes a closer look at the causes of death among homeless
people, focusing specifically on deaths related to drug addiction. Using
data visualization, the paper identifies drug toxicity as the leading
cause of death among homeless people. @fig-causes-of-death-by-age
provides a breakdown of causes of death, highlighting the proportion
attributable to addiction-related problems. These deaths are an example
of the negative health impacts of the ongoing drug toxicity crisis. This
section discusses the implications of these findings for public health
initiatives and recommendations for addressing addiction within the
homeless community. Over the past several years, Toronto Public Health,
the City of Toronto and community partners have used ongoing funding for
harm reduction services and supports and affordable housing. These
policies can really help help homeless people with drug addictions and
other homeless people get safer housing and adequate medical support.
From a medical safety perspective, these policies may even help homeless
drug addicts gain access to safer and more stable places to use drugs.
These policies help reduce the number of deaths among drug addicts.

## The rising number of homeless deaths caused by drugs:

The data analysis also revealed a disturbing trend of rising ratios of
drug toxicity deaths among homeless people in Toronto. This section
discusses the reasons for this upward trajectory, emphasizing the
urgency for targeted interventions. By studying the annual distribution
of deaths, we reveal patterns and potential influencing factors that may
inform policy changes and resource allocation. @fig-ratio-of-death shows,
drug Toxicity has become the main cause of death among homeless people
since 2017. Before the COVID-19 pandemic, the death ratio among homeless
people due to drug toxicity remained at around 40%. And after the
pandemic began in 2020, the ratio of homeless deaths caused by drug
toxicity quickly increased to 65%, a 26% increase in one year. And it
maintained this ratio in the following years, and jumped again to 73% in
2023, an increase of 10%. The ratio of drug toxicity deaths among
homeless people in Toronto continues to rise since 2020. This was also
the time of year when the Covid 19 pandemic began. Dr. Stephen Hwang, a
physician and researcher at the MAP Center for Urban Health Solutions at
Allied Health in Toronto, said the disruption caused by lock-downs and a
lack of services has led people to change where they spend their time,
and therefore where they use drugs (CityNews Toronto, 2023). Due to the
impact of the pandemic, Toronto Public Health and the City of Toronto
have had to significantly increase their funding expenditures in
response to the pandemic, thereby reducing some support services for the
homeless population. The impact of lock-downs and work suspensions has
also led to a crisis in the lives and health of homeless people.

# Result

Our analysis presents a portrayal of the escalating impact of
drug-related deaths within Toronto's homeless population. The confluence
of drug addiction and the COVID-19 pandemic has resulted in a profound
surge. Effectively addressing this challenge demands not only sustained
funding for harm reduction services but also responses tailored to the
challenges posed by the pandemic. A comprehensive understanding of the
dynamics of causes and patterns of death becomes paramount as we
confront these complex issues. This knowledge is instrumental in shaping
policies that can genuinely save lives within the vulnerable homeless
communities. The findings underscore the critical importance of drug
safety outreach specifically tailored to homeless drug users. It becomes
imperative to extend harm reduction and treatment options directly to
those on the streets or within the community, rather than relying solely
on clinic or hospital-based approaches. Waiting for individuals to seek
help within traditional healthcare settings may prove inadequate in
addressing the urgent needs of this vulnerable population. Proactive and
accessible interventions are essential to mitigating the devastating
impact of drug-related deaths among the homeless. This shift towards
community-based outreach aligns with the recognition that meeting
individuals where they are can significantly enhance the effectiveness
of harm reduction and treatment initiatives.

## References
---
nocite: '@*'
---
```{r}
#| warning: false
#| message: false
#| echo: false
knitr::write_bib(c("base","knitr", "rmarkdown","janitor","lubridate","tidyverse","opendatatoronto","dplyr"), "references.bib")
new_entry <- "@misc{Opioid_overdoses_among_homeless_people_rose_dramatically_2023,
  title = {Opioid overdoses among homeless people rose dramatically: Ontario study},
  author = {},
  howpublished = {\\url{https://toronto.citynews.ca/2023/10/17/opioid-overdose-deaths-homeless-ontario/}},
  journal = {CityNews Toronto},
  year = {2023},
  month = {Oct},
  note = {Retrieved from \\url{https://toronto.citynews.ca/2023/10/17/opioid-overdose-deaths-homeless-ontario/}}
}"

# Append the new entry to the "references.bib" file
write(new_entry, "references.bib", append = TRUE)

@@
